---
title: "Bush stone-curlew behaviour and performance analyes"
author: "Shoshana Rapley"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)

# Load packages
pacman::p_load(contsurvplot, ggpubr, glmmTMB, janitor, modelsummary, MuMIn, patchwork, performance, readxl, sjPlot, survival, survminer, tidyverse)
```

# Introduction

This Markdown documents analyses done as part of paper 'Behavioural predictors of survival change over time for a translocated bird', a chapter of my doctoral studies. 

The aim of this study is to test whether we can predict post-release performance using pre-release assays in translocated bush stone-curlews. If true, we might be able screen for suitable release candidates in future translocations. 

The purpose of the translocation was to hold the birds in a quasi-wild setting as an extension of the captive breeding facility. Birds from Mt Rothwell were destined for reintroduction to Orana Sanctuary in 2023.

In this study we translocated 16 bush stone-curlews into the Mt Rothwell fenced sanctuary from the Mt Rothwell breeding colony. They were released in October 2022 to January 2023 across three release groups. This was the pilot release at the site (i.e., no conspecifics prior to the translocation and a high degree of risk and uncertainty), which is why we released in a 'drip feed' manner over the Summer of 2022/23. 

Two individuals were excluded from analyses. Daisy was returned to captivity after two days due to risky behaviour (hanging around the buildings in a high traffic area). Star died of gastroenteritis shortly after release and so her fate was not linked to behaviour. 

# Predictors

We undertook pre-release assays prior to release of the birds. The pre-release assays were:

a) Latency to reach food
b) Handling response (in-hand)
c) Filmed behaviour

Predictors A and C were filmed in their home aviaries with trail cameras. Predictor B was scored during routine handling.

## Latency

Latency to reach food was defined as how long it took each bird, in seconds, to reach a plate of food provided as part of their normal husbandry. It was measured from when the plate was placed on the ground until the bird took the first mouthful. 

```{r latency data}
# Read in and clean up latency assay data
latency <- read_excel("Input/bsc traits and assay data.xlsx", sheet = "Latency") %>%
  clean_names() %>%
  # Format dates
  mutate(date = as_date(date)) %>%
  # select columns
  dplyr::select(c("date", "identity", "reach", "latency_s")) %>%
  # Convert yes/no to 1/0
  mutate(reach = ifelse(reach == "Yes", 1, 0)) %>%
  rename(latency = latency_s) %>%
  # Add column for plotting 'did not reach'
  mutate(label = ifelse(is.na(latency), "DNR", ""))

# Plot latency by bird
ggplot(latency) + 
  geom_bar(aes(identity, latency, fill = as.character(date)), 
           position = "dodge", stat = "identity")+
  geom_text(aes(y=600, x= identity, label = label), angle = 90, vjust = -0.5)+
  labs(x= element_blank(), y = "Latency to reach food (seconds)")+
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  scale_fill_manual(values = c("#881C00FF","#1BB6AFFF"), name = "Assay date")+
  theme_minimal()+
  ggplot2::theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

We need to account for repeat measures but due to low sample size we cannot use bird as a random factor (model convergence issues when tested). So instead, we are going to use the mean latency. The birds didn't reach the food on one third of occasions, but all birds reached the food at least once (out of two trials).

```{r mean latency}
# Calculate mean latency by bird
mlatency <- latency %>% 
  na.omit() %>%
  group_by(identity) %>%
  summarise(latency = mean(latency))

# Plot mean latency by bird
ggplot(mlatency) + 
  geom_bar(aes(identity, latency), position = "dodge", stat = "identity")+
  labs(x= element_blank(), y = "Mean latency to reach food (seconds)")+
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  scale_fill_manual(values = c("#881C00FF","#1BB6AFFF"), name = "Assay date")+
  theme_minimal()+
  ggplot2::theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

# Check the distribution of mean latency
hist(mlatency$latency)
shapiro.test(mlatency$latency)
qqnorm(mlatency$latency)

# Transform mean latency
hist(sqrt(mlatency$latency))
shapiro.test(sqrt(mlatency$latency))
qqnorm(sqrt(mlatency$latency))
```

I applied a square-root transformation to latency because the data was left-skewed. The Shapiro-Wilks test for normalcy showed the pre-transformed data were non-normally distributed while the post-corrected data were (better, but still imperfectly) normally distributed. 

## Handling response

Handling response was assigned during processing, and is an ordinal variable on a scale of 1:3:

1) Calm/still/log-pose
2) Moderate/slight kicking or struggling/soft calling or beak clacking
3) Extreme/upset/shrieking/growling

```{r handling}
# Read in handling data
handling <- read_excel("Input/bsc traits and assay data.xlsx", sheet = "Ratings") %>%
  clean_names() %>%
  # Format dates
  mutate(date = as_date(date)) %>%
  # Drop unneeded columns
  dplyr::select(-c(date, notes))

# Plot frequency of handling scores
ggplot(handling)+
  geom_bar(aes(score))+
  scale_y_continuous(breaks =c(1,3,5,7,9))+
  theme_minimal()+
  ggplot2::theme(panel.grid.minor = element_blank())
```

Most birds (64%) had a handling score of 1, with 21% a score of 2 and just two individuals (14%) with a score of 3. 

As a side note, where we have repeat measures (i.e., taken post-release- but these aren't documented here) the handling response stays the same; however, for now we are just interested in whether the pre-release assay values predict other post-release behaviours. 

## Behaviour

Birds were filmed in their aviaries between 12pm-4pm. For every film recorded (activated by the infrared sensor on the trail camera) we encoded each second of behaviour for each bird. We calculated the the proportion of time spent in alert behaviours (considered mutually exclusive from 'relaxed' behaviours). 

*Alert behaviours* were defined as:

* Alertness: standing but not relaxed, often startled into this pose from a more relaxed posture. May have a stiffer pose, holding neck extended or forward. May have flattened the feathers against the body. 
* Postured walk: walking with the neck stiffly extended parallel wit the ground. Feathers either fluffed up (held erect) or slimmed against body, as opposed to held loosely. 
* Head-bobbing: a rapid up/down movement of the head; used to communicate unease.
* Head-tilting: craning the head to the side to look above and around; possibly to check for danger (can be done in tandem with head-bobbing).
* Tail-wagging: a rapid side/side shake of the tail; used to communicate unease.
* Defensive display: holding the wings outstretched; often wings are rapidly extended which gives the impression of a black and white 'flash' of plumage.

*Relaxed behaviours* were defined as:

* Preening: actively cleaning self (i.e., running the beak along the feathers, rearranging plumage), or scratching (e.g. the face with the toes), or stretching (e.g. extending the leg and wing).
* Fluffing: a rapid erection and flattening of the feathers, often accompanied by a body and head shake. Often preceded by preening. 
* Standing: relaxed pose (in the absence of vigilant behaviours as defined above), feathers loose or lightly fluffed (as opposed to erect or flat against the body).
* Walking: stepping forward (in the absence of vigilant behaviours as defined above).
* sitting: either on tibiotarsus or fully on the ground; includes sleeping. 
* Eating: engaged in looking at, carrying, pecking, picking up, swallowing or breaking up food (or non food items that are being tested e.g. leaves). Can rapidly (second by second) change between eating and alertness. 
* Calling: calling as part of group crescendo with "werloo" song (as opposed to short contact calls while doing other beahviours such as sitting or standing)

Here we read in the video assay data and calculate the time spent vigilant as a proportion of whole time. Some videos have sections where the bird is out of frame (noted in ethogram as 'Z'), which are excluded from the time allocation analysis. 

```{r behaviours}
# Ethogram
eth <- matrix(c("A", "Alert",
                "C", "Calling",
                "D", "Defensive wing display",
                "E", "Eating/drinking",
                "F", "Fluff",
                "HB", "Head bob",
                "HT", "Head tilt",
                "I", "Sitting",
                "N", "Camouflaged walk",
                "O", "Courtship",
                "P", "Preening",
                "S", "Standing",
                "W", "Walking",
                "TW", "Tail wag",
                "Z", "Out of frame"),
              ncol = 2, byrow = TRUE) %>%
  as.data.frame() %>%
  rename(code = V1,
         Behaviour = V2)

# Assign plotting order of behaviours
order <- c("Alert", "Head bob", "Head tilt", "Tail wag", "Camouflaged walk", 
           "Defensive wing display", "Calling", "Eating/drinking", "Courtship",
           "Fluff", "Preening", "Sitting", "Standing", "Walking", "Out of frame")

# Read in and tidy up video assays 
behaviour <- read_excel("Input/bsc traits and assay data.xlsx", sheet = "Video encoding") %>%
  clean_names() %>%
  # Drop metadata columns
  dplyr::select(-c("clip", "note")) %>%
  # Convert from wide to long
  pivot_longer(3:61, names_to = "second", values_to = "code") %>%
  # Convert seconds into numeric
  mutate(second = as.numeric(str_remove(second, "x"))) %>%
  # Drop instances of out of frame 
  filter(!code == "Z") %>%
  # Lump all variations of head tilt 
  mutate(code = ifelse(code %in% c("HBT", "HTB", "HTW"), "HT", code)) %>%
  # Add behaviour descriptions from ethogram
  left_join(eth) %>%
  # Turn behaviour into a factor to assign order for plotting
  mutate(Behaviour = factor(Behaviour, levels = order)) %>%
  # Assign vigilance to 1 and not to 0
  mutate(vigilant = ifelse(code %in% 
                             c("A", "D", "HB", "HT", "TW", "N"), 1, 0)) %>%
  # Drop lines without recordings (primarily for Prem and Rove who had 10 second clips)
  na.omit()

# Plot proportion of time spent in each behaviour per bird
ggplot(behaviour) + 
  geom_bar(aes(x= identity, fill = Behaviour), position = "fill")+
  labs(x= element_blank(), 
       y = "Proportion of time spent in each behaviour")+
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  theme_minimal()+
  ggplot2::theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank())+
  scale_fill_viridis_d(direction = -1)

# save to file
ggsave("plots/time_spent_behaviours.png", dpi = 400, width = 20, height = 15, units = "cm", bg = 'white')

# Calculate proportion of time spent in each behaviour and a summed 'alert' metric
time <- count(behaviour, identity) %>%
  rename(total = n)

prop_behav <- count(behaviour, identity, code) %>%
  left_join(time) %>%
  # Divide count of behaviour by total time for %
  mutate(prop = n/total) %>%
  dplyr::select(-c(n, total)) %>%
  # Pivot wider, one column per behaviour
  pivot_wider(names_from = code, values_from = prop) %>%
  # Replace NAs for behaviour with zeroes
  replace(is.na(.), 0) %>%
  # Add column for summed 'alert' measure
  mutate(alert = A + HB + HT + TW + N + D)

# Plot time spent alert
ggplot(prop_behav) +
  geom_bar(aes(x= identity, y= alert), position = "dodge", stat = "identity")+
  labs(x= element_blank(), 
       y = "Proportion of time spent in alert behaviours")+
  scale_x_discrete(guide = guide_axis(angle = 45)) +
  theme_minimal()+
  ggplot2::theme(panel.grid.minor = element_blank())

# Check distribution of behaviours
hist(prop_behav$alert)

# Summarise allocation of behaviours
prop_behav_sum <- prop_behav %>%
  summarise(across(where(is.numeric),list(mean = mean, sd = sd))) %>%
  pivot_longer(1:30)
```

Interestingly, the amount of time spent 'alert' (yellows and greens) seems (visually at least) inversely related to the amount of time spent eating (aqua), rather than cutting in to the time spent resting (purples). 

The 'alert' metric is close to normally distributed. 

## Traits

I included some non-behavioural trait predictors to compare with the assays predictions. These included life-history and morphological traits.

```{r traits}
# Sex
sex <- read_excel("Input/bsc traits and assay data.xlsx", sheet = "Metadata") %>%
  clean_names() %>%
  select(c(identity, sex))

# Pre-release weights
weights <- read_excel("Input/bsc traits and assay data.xlsx", sheet = "Weights") %>%
  clean_names() %>%
  # drop unneeded date column
  dplyr::select(-date)

# combine
traits <- left_join(sex, weights)

# Summary of weights
weight_sum <- traits %>%
  group_by(sex) %>%
  summarise(weight_mean = mean(weight),
            weight_sd = sd(weight))
```

Out of curiosity, I want to know if weight and sex are correlated.

```{r traits 2}
# Weight by sex
ggplot(traits, aes(x = sex, y = weight))+
  geom_boxplot()+
  theme_minimal()
```

Males weigh more than females.

# Responses

We tracked birds with GPS devices to determine their fate. The GPS devices were removed in January 2024. Because different cohorts were released at different times (October, December or January), the maximum tracked duration is 363 days (which we call a year, close enough)

We assessed survival in three ways:

* survival to 90-days post-release (yes/no; binary)
* survival to 1-year post-release (yes/no; binary)
* persistence (how many days survived; continuous)

We tested survival to both 90-days and 1-year post-release becasue we wanted to know whether the time period of monitoring impacted the strength of prediction. We chose 90-days because this was the typical tracking duration in previous bush stone-curlew translocation when tracking was done with VHF (which was at Mulligans Flat in 2014-2016).

```{r survival}
# Read in survival data
survival <- read_excel("Input/bsc traits and assay data.xlsx", sheet = "Survival") %>%
  clean_names() %>%
  # Limit persistence value to minimum common tracked time of 360
  mutate(persist = ifelse(persist>363, 363, persist)) %>%
  # Convert true and false to 0 and 1
  mutate(to_90 = ifelse(to_90 == FALSE, 0, 1),
         to_363 = ifelse(to_363 == FALSE, 0, 1)) %>%
# Drop unneeded columns
  dplyr::select(c(identity, persist, to_90, to_363))

# Combine all traits and response (survival)
data <- left_join(survival, mlatency) %>%
  left_join(handling) %>%
  left_join(prop_behav) %>%
  left_join(traits) %>%
  # Drop un-needed behaviour columns
  dplyr::select(-c(A, C, E, F, HB, HT, N, O, P, S, TW, W, D, I))

# Write table to file
write.csv(data, row.names = FALSE, "Output/data_table.csv")
```

# Analyses

```{r}
# read in data
data <- read_csv("Output/data_table.csv") %>%
  # drop Star and Daisy
  filter(!identity %in% c("Daisy", "Star"))
```

## Assay correlation

Are these metrics related? I.e. are there consistency within individuals in the types of responses across the different measures. 

```{r correlated assay}
m0 <- glm(score ~ alert * latency,
          data = data)

summary(m0)
```

Assays seem to be unrelated - so are measuring different things. 

## Binary survival

First I take a hypothesis driven approach to looking at the behaviours and traits, to see which predict survival to 90 days and 1 year. I expect to see a difference in which predict survival for each time period, and I expect weight and food to be more important in the short term. I don't expect to see a difference in traits other than pre-release weight. 

### m1 Alert behaviour and survival

```{r m1}
# m1a: Does proportion of time spent alert predict survival to 90 days?
## Violin plot
ggplot(data, aes(x= to_90, y = alert, group=to_90))+
  geom_violin()+
  geom_point(size = 2)+
  labs(x= "Survival to 90-days post-release", 
       y = "Proportion of time spent alert")+
  theme_minimal()+
  ggplot2::theme(panel.grid.minor = element_blank())+
  scale_x_continuous(breaks = c(0,1), labels = c(FALSE, TRUE))

## Logistic regression with binomial family
m1a <- glm(to_90 ~ scale(alert),
           data = data,
           family = binomial())

summary(m1a)

# m1b: Does proportion of time spent alert predict survival to 1 year?
## Violin plot
ggplot(data, aes(x= to_363, y = alert, group=to_363))+
  geom_violin()+
  geom_point(size = 2)+
  labs(x= "Survival to 1 year post-release", 
       y = "Proportion of time spent alert")+
  theme_minimal()+
  theme(panel.grid.minor = element_blank())+
  scale_x_continuous(breaks = c(0,1), labels = c(FALSE, TRUE))

## Logistic regression with binomial family
m1b <- glm(to_363 ~ scale(alert),
           data = data,
           family = binomial())

summary(m1b)
```

No effect of behaviour on survival to either time period. Which on the one hand is annoying (because it took 40 hours to code the data) but on the other hand is useful (because it might save another practitioner doing the same). It also might be the wrong metric. The birds spent half of their time asleep, so perhaps the behaviour metrics should be recorded at night when the birds are active instead of during their period of restfulness. 

### m2 Latency and survival

```{r m2a}
# m2a: Does latency to reach food predict survival to 90 days?
## Violin plot
ggplot(data, aes(x= to_90, y =sqrt(latency), group=to_90))+
  geom_violin()+
  geom_point(size = 2)+
  labs(x= "Survival to 90-days post-release", 
       y = "Latency to reach food (squareroot transformed)")+
  theme_minimal()+
  theme(panel.grid.minor = element_blank())+
  scale_x_continuous(breaks = c(0,1), labels = c(FALSE, TRUE))

## Logistic regression with binomial family
m2a <- glm(to_90 ~ scale(sqrt(latency)),
           data = data,
           family = binomial())

summary(m2a)

# Survival probability
exp(coef(m2a))
```

Only three birds (21%) died in the period to 90-days post-release. There is weak evidence (i.e. marginally significant prediction) that shorter latency periods predict survival. 

When latency approaches zero the chance of survival is 82% (based on the intercept coefficient estimate). With every unit increase in sqrt(latency), the odds of survival decrease by a factor of 0.94.

Model performance is imperfect, with non-normally distributed residuals. Given the management needs to balance risk and uncertainty during translocation, I would use this imperfect information to inform the translocation approach until better information can be gathered for adaptive management. Since lower latency *might* inform survival, and we want to reduce unnecessary mortality, this is a useful management tool. 

Next, we looked at the effect of latency on survival to 1-year post-release. 

```{r m2b}
# m2b: Does latency to reach food predict survival? To 1 year.
## Violin plot
ggplot(data, aes(x= to_363, y=sqrt(latency), group=to_363))+
  geom_violin()+
  geom_point(size = 2)+
  labs(x= "Survival to 1-year post-release", 
       y = "Latency to reach food (squareroot transformed)")+
  theme_minimal()+
  theme(panel.grid.minor = element_blank())+
  scale_x_continuous(breaks = c(0,1), labels = c(FALSE, TRUE))

## Logistic regression with binomial family
m2b <- glm(to_363 ~ scale(sqrt(latency)),
           data = data,
           family = binomial())

summary(m2b)
```

Just under half of the birds (42%, n = 6) survived to the 1-year post-release period. The chance of survival increases with decreasing latency values. However, unlike for 90-days, the effect is not significant. This could indicate that latency has a more immediate impact on short-term survival but less influence on long-term survival.

Potentially, whatever behavioural traits are related to latency (perhaps risk:reward trade-off or drive for food) only relates to survival in the period immediately following release. Perhaps this is due to the drive to find food resources during establishment, but this effect does not impact long-term survival which is (likely) more related to anti-predation traits, rather than resource-related traits.

This highlights the importance of sampling period and the different drives of survival in the short and long term post-release.

In summary, latency only has a meaningful impact on short-term survival (to 90 days), not overall persistence across the 1-year study period. 

### m3 Handling response and survival

Since handling score is categorical, we'll use ANOVA and chi-squared to compare frequencies. 

```{r m3}
# m3a: Does handling predict survival to 90 days?
## Mosaic plot
mosaicplot(table(data$to_90, data$score), 
           main = NULL, 
           xlab = "Survival to 90-days post-release",
           ylab = "Handling score")

## Chisq test
m3a <- chisq.test(table(data$to_90, data$score))

m3a

# m3b: Does handling predict survival to 1 year?
## Mosaic plot
mosaicplot(table(data$to_363, data$score), 
           main = NULL, 
           xlab = "Survival to 1-year post-release",
           ylab = "Handling score")

## Chisq test
m3b <- chisq.test(table(data$to_363, data$score))

m3b
```

While there is no evidence of a relationship between handling score and survival to 90-days, there is weak evidence of a relationship to 1-year.

Some interesting observations:

1) Both birds with a handling score of 3 survived. 
2) No birds with a handling score of 2 survived. 

Perhaps an intermediate strategy is sub-optimal i.e., it's better to either commit to full camouflage or full flight, not the middle ground. Perhaps the middle ground is evidence of habituation, since both scores 1 and 3 can be considered a predator-response behaviour (flight or camouflage). 

While latency to reach food better predicted survival to 90-days, handling response better predicted survival to 1-year. As discussed earlier, perhaps that is because these assay approaches measure different traits. I think latency may be related to resource acquisition while handling may be related to anti-predator response. 

Interestingly, the persistence for score 2 individuals is pretty long (195 & 308 days) - which contradicts my earlier thoughts about score 2 being maladaptive. 

Overall, there is weak evidence that handling score predicts persistence. In the context of adaptive management (the same as for the latency score), I would use this information to inform management until better information can be gathered. That is, preferentially release score 3 birds when risk and uncertainty are high. I would not release score 2 birds during a pilot study, but perhaps they can be used later in a reintroduction program. 

In summary, handling score is a better predictor of long-term surival (1 year) than latency. 

### m4 Weight and survival

```{r m4}
# m4a: Does weight predict survival to 90 days?
## Violin plot
ggplot(data, aes(x= to_90, y=weight, group=to_90))+
  geom_violin()+
  geom_point(size = 2)+
  labs(x= "Survival to 90-days post-release", 
       y = "Pre-release weight")+
  theme_minimal()+
  theme(panel.grid.minor = element_blank())+
  scale_x_continuous(breaks = c(0,1), labels = c(FALSE, TRUE))

## Logistic regression with binomial family
m4a <- glm(to_90 ~ scale(weight),
           data = data,
           family = binomial())
summary(m4a)

# m4b: Does weight predict survival to 1 year?
## Violin plot
ggplot(data, aes(x= to_363, y=weight, group=to_363))+
  geom_violin()+
  geom_point(size = 2)+
  labs(x= "Survival to 1-year post-release", 
       y = "Pre-release weight")+
  theme_minimal()+
  theme(panel.grid.minor = element_blank())+
  scale_x_continuous(breaks = c(0,1), labels = c(FALSE, TRUE))

## Logistic regression with binomial family
m4b <- glm(to_363 ~ scale(weight),
           data = data,
           family = binomial())
summary(m4b)
```

No effect of weight on survival. 

### m5 Sex and survival

A predictor for survival to 1-year was sex I investigate this separately here. 

```{r m5}
# m5a: Does sex predict survival? To 90 days.
## Mosaic plot
mosaicplot(table(data$to_90, data$sex), 
           main = NULL, 
           xlab = "Survival to 90-days post-release",
           ylab = "Sex")

chisq.test(table(data$to_90, data$sex))

## Logistic regression with binomial family
m5a <- glm(to_90 ~ sex,
           data = data,
           family = binomial())

summary(m5a)

# m5b: Does sex predict survival? To 1 year.
## Mosaic plot
mosaicplot(table(data$to_363, data$sex), 
           main = NULL, 
           xlab = "Survival to 1-year post-release",
           ylab = "Sex")

chisq.test(table(data$to_363, data$sex))

## Logistic regression with binomial family
m5b <- glm(to_363 ~ sex,
           data = data,
           family = binomial())

summary(m5b)
```

Sex did not predict survival in either time period. The issue is there are only 2 males not surviving to 1 year and none not surviving to 90 days. There isn't enough predictive power. But it is interesting to mention. I wonder if sex is coming up as important because the males are bigger and bigger birds are surviving better? 

## Survival analysis

Instead of looking at survival as a binary yes/no, we can also look at persistence as a continuous response. For this, I am using a survival analysis approach, since 'persistence' is a time-to-event response. 

Note: cannot test handling response due to model convergence issues (too few members per group).

```{r}
# Format data for Kaplan Meier Analysis - where death is TRUE i.e. 1
km <- data %>%
  mutate(time = persist,
         status = ifelse(to_363 == 0, 1, 0),
         sex = as.factor(sex),
         score = as.factor(score)) %>%
  select(-c(persist, to_90, to_363))
```

### m6 Survival (all)

Start by looking at the overall survival probabilities for all birds. 

```{r m6}
# Fit a survival model
m6 <- survfit(Surv(time, status) ~ 1, km)

# Estimated survival probabilities
summary(m6, times = c(1,90, 180, 270, 363))
```

The probability of survival to 1 year is 43% (upper CI = 78%, lower CI = 23%). 

### m7 Latency and persistence 

```{r m10}
# Fit a survival model with latency as the continuous variable
m7 <- coxph(Surv(time, status) ~ sqrt(latency), km, x= TRUE)

# Plot the survival area
plot_surv_area(model = m7, data = km,
               time = "time", status = "status", variable = "latency", group = NULL)

# Test effect of latency - log rank test
summary(m7)
```

Evidence for effect of latency, where shorter latency periods have higher survival likelihoods.

### m8 Alert behaviour and persistence 

```{r m11}
# Fit a survival model with alert behaviour as the continuous variable
m8 <- coxph(Surv(time, status) ~ alert, km, x= TRUE)

# Plot the survival area
plot_surv_area(model = m8, data = km,
               time = "time", status = "status", variable = "alert", group = NULL)

# Test effect of alert behaviour - log rank test
summary(m8)
```

No effect of alertness on survival.

### m9 Sex, weight and persistence 

```{r m12}
# Weight
m9a <- coxph(Surv(time, status) ~ weight, km, x= TRUE)
summary(m9a)

# Sex
m9b <- coxph(Surv(time, status) ~ sex, km)
summary(m9b)
```

Weight and sex alone do not explain persistence. However, visually, it looks like bigger females survive better while the opposite is true for males. So it might be better to combine weight * sex. 

```{r}
m9c <- coxph(Surv(time, status) ~ weight * sex, km, x= TRUE)

# Plot the survival area
plot_surv_area(model = m9c, data = km,
               time = "time", status = "status", variable = "weight", group = "sex")

# Test effect of weight - log rank test
summary(m9c)
```

Combined effects significant.

# Model effects summary tables

```{r}
# binary response effects - not including hanlding which is chi squared instead of glm
modelsummary(list(m1a, m1b, m2a, m2b, m4a, m4b, m5a, m5b), 
             estimate = "{estimate}", statistic = c("{std.error}", "{p.value}"), coef_omit = "Intercept",
             output = "Output/model_summary_binary.csv")

# time-to-event model effects
modelsummary(list(m7, m8, m9a, m9b, m9c), 
             estimate = "{estimate}", statistic = c("{std.error}", "{p.value}"), coef_omit = "Intercept",
             output = "Output/model_summary_time_to_event.csv")
```

# Figures

Making some nicer looking figures for publication here. 

```{r}
# read in data
data <- read_csv("Output/data_table.csv", show_col_types = FALSE) %>%
  # drop Star and Daisy
  filter(!identity %in% c("Daisy", "Star")) %>%
  # rename period labels to look nicer on graphs
  rename("90 days" = to_90,
         "1 year" = to_363) %>%
  # pivot wider on period
  pivot_longer(3:4, names_to = "period", values_to = "survived") %>%
  rename(Survival = survived) %>%
  # categorical variables as factors
  mutate(period = as_factor(period),
         score = as_factor(score),
         Survival = factor(Survival, levels = c("1", "0"))) 

# and separately format data for survival analysis
data2 <- data <- read_csv("Output/data_table.csv", show_col_types = FALSE) %>%
  # drop Star and Daisy
  filter(!identity %in% c("Daisy", "Star")) %>%
  mutate(time = persist,
         status = ifelse(to_363 == 0, 1, 0),
         sex = factor(sex),
         score = factor(score)) %>%
  select(-c(persist, to_90, to_363))
```

The binary survival yes/no to 90 days and 1 year for all variables (continuous as violin and categorical as mosaic). 

```{r}
# default theme options for all plots
theme <- function(){
  list(
    geom_violin(),
    geom_point(),
    facet_wrap(~period),
    xlab(element_blank()),
    theme_classic(),
    scale_x_discrete(labels = c("Survived", "Died")),
    ggplot2::theme(axis.title.y = element_text(size = 11))
  )
}

# list of plots
g1 <- ggplot(data, aes(Survival, latency))+
  ylab("Latency to reach food (s)")+
  theme()

g2 <- ggplot(data, aes(Survival, alert))+
  ylab("Time spent alert %")+
  theme()

g3 <- ggplot(data, aes(Survival, weight))+
  ylab("Pre-release weight (g)")+
  theme()

g4 <- ggplot(data, aes(x = Survival, y = weight, color = sex)) +
  ylab("Pre-release weight (g)") +
  geom_violin(aes(group = interaction(Survival, sex, period)), 
              position = position_dodge(width = 0.9)) +
  geom_point(aes(group = interaction(Survival, sex, period)), 
             position = position_dodge(width = 0.9)) +
  facet_wrap(~period) +
  scale_colour_manual(values = c("#298c8c", "#f1a226"),
                      name = "Sex")+
  xlab(element_blank()) +
  theme_classic()+
  scale_x_discrete(labels = c("Survived", "Died"))+
  ggplot2::theme(axis.title.y = element_text(size = 11))

g5 <- ggplot(data)+
  geom_bar(aes(x = score, fill = Survival), position="fill")+
  theme_classic() +
  scale_fill_manual(values = c("gray", "gray5"), 
                    labels = c("Survived", "Died"),
                    name = "Survival") +
  facet_wrap(~period) +
  xlab("Handling score") +
  ylab("Survival proportion")+
  ggplot2::theme(axis.title.y = element_text(size = 11))

g6 <- ggplot(data)+
  geom_bar(aes(x = sex, fill = Survival), position = "fill")+
  theme_classic() +
  scale_fill_manual(values = c("gray", "gray5"), 
                    labels = c("Survived", "Died"),
                    name = "Survival") +
  facet_wrap(~period) +
  xlab("Sex") +
  ylab("Survival proportion")+
  ggplot2::theme(axis.title.y = element_text(size = 11))

# arrange multi plot - moved from ggarrange to patchwork, seems to work better
combined_plot <- (g1 + g2) / (g3 + g4) / (g5 + g6) + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A')
combined_plot

# save
ggsave("plots/binary_response_figure.png", dpi = 400, width = 22, height = 20, units = "cm")
```

Second, the continuous persistence variables

```{r}
s1 <- survminer::ggsurvplot(survfit(Surv(time, status) ~ sex, data2),
                 conf.int = TRUE,
           legend.title = "Sex",
           legend = "right",
           palette = c("#298c8c", "#f1a226"),
           legend.labs = c("Female", "Male"),
    xlab = "Days post release",
    font.x =  12,
    font.y = 12,
    font.tickslab = 10,
    ggtheme = theme_bw(),
    xlim = c(0, 365),
    break.time.by = 100,
    censor = FALSE)

s2 <-  contsurvplot::plot_surv_area(model = coxph(Surv(time, status) ~ weight, data2, x= TRUE),
               status = "status",
               variable = "weight",
               time = "time",
               data = data2,
               group = NULL,
               legend.title = "Pre release\nweight (g)",
               start_color="gray",
               end_color="black",
               alpha = 0.8,
    xlab = "Days post release",
    font.x =  12,
    font.y = 12,
    font.tickslab = 10,
    ggtheme = theme_bw(),
    xlim = c(0, 365),
    break.time.by = 100,
    censor = FALSE)

s3 <- contsurvplot::plot_surv_area(model = coxph(Surv(time, status) ~ sex * weight, data2, x= TRUE),
               data = data2,
               time = "time", 
               status = "status", 
               variable = "weight", 
               group = "sex",
               start_color="gray",
               end_color="black",
               legend.title = "Pre release\nweight (g)",
    xlab = "Days post release",
    font.x =  12,
    font.y = 12,
    font.tickslab = 10,
    ggtheme = theme_bw(),
    xlim = c(0, 365),
    break.time.by = 100,
    censor = FALSE)

# arrange multi plot - moved from ggarrange to patchwork, seems to work better
combined_plot <- (s1$plot + s2) / (s3)+ 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = 'A')
  
combined_plot

# save
ggsave("plots/weight_sex_persistence.png", dpi = 400, width = 20, height = 14, units = "cm")

# and latency by itself
s4 <-  contsurvplot::plot_surv_area(model = coxph(Surv(time, status) ~ latency, data2, x = TRUE),
               status = "status",
               variable = "latency",
               time = "time",
               data = data2,
               group = NULL,
               legend.title = "Latency to\nreach food\n(seconds)",
               start_color="gray",
               end_color="black",
               alpha = 0.8,
    xlab = "Days post release",
    font.x =  12,
    font.y = 12,
    font.tickslab = 10,
    ggtheme = theme_bw(),
    xlim = c(0, 365),
    break.time.by = 100,
    censor = FALSE)
s4

# save
ggsave("plots/latency_persistence.png", dpi = 400, width = 14, height = 10, units = "cm")
```
