---
title: "p2_assays"
author: "Shoshana Rapley"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, janitor, lubridate, SDLfilter, ggmap, amt,
               gganimate, sp, terra, adehabitatHR, sf)

ggmap::register_google(key = "AIzaSyAH3qnqmxDATEluG8lKt2KtnHKLYJc2WaM")

```

# Introduction

This Markdown documents pre-release assays and post-release survival and establishment metrics for translocated
bush stone-curlews (curlews) at Mt Rothwell, Victoria. 

# Post-release data

Data are stored on Movebank, and were downloaded for the period 25/10/2022 to 10/05/2023. Data are cleaned to 
remove erroneous GPS fixes. 

```{r data clean}

birds <- c("Briar", "Nutmeg", "Star", "Aurora", "Robin", "Daisy",
           "Iona", "Sage", "Koda", "Brook", "Clover", "Wobbles",
           "Prem", "Rove", "Valentine", "Marmalade")

# import data from movebank and filter to study birds 
data_raw <- read.csv("movebank_20230510.csv") %>%
  clean_names() %>%
  rename(id = individual_local_identifier,
         lon = location_long,
         lat = location_lat,
         qi = gps_satellite_count) %>%
  filter(id %in% birds) %>%
  mutate(DateTime = as.POSIXct(study_local_timestamp,
                               "%Y-%m-%d %H:%M:%S",
                               tz = "Australia/Sydney")) %>%
  mutate(date = date(DateTime))

# filter out points with a hdop >1.4
data_hdop <- data_raw %>%
  filter(qi>5,
         gps_hdop<1.4) #hdop recent addition 16 jan to improve accuracy

# filter out points using speed-based filtering
data_filt <- data_hdop %>%
  dupfilter() %>%
  ddfilter_speed(vmax=4, method=2)%>%
  rename(date_time = DateTime)

# plot data cleaning to check visually
map <- get_map(c(144.435738, -37.900544), zoom=14, maptype = "satellite")
ggmap(map)+
  geom_point(data=data_raw, aes(lon, lat), color="red")+
  geom_point(data=data_hdop, aes(lon, lat), color="yellow")+
  geom_point(data=data_filt, aes(lon, lat), color="blue")

# save data
write.csv(data_filt, "movebank_20230510_clean.csv", row.names = FALSE)

```

## Distance

Here we calculate two distance metrics: 
  1) distance from release location
  2) total distance moved daily

```{r distance from release}

# read in clean data
data <- read.csv("movebank_20230510_clean.csv")

# specify the coords of each GPS fix
coords <- as.matrix(cbind(data$utm_easting, data$utm_northing))

# set coords for the release location 
release  <- matrix(c(274423.45, 5801912.85), ncol=2) #(-37.902374, 144.434323)

# calculate distance between the release location and all fixes
dist <- mutate(data, dist_release = distance(coords, as.matrix(release), lonlat=FALSE))

# calculate daily summary statistics
dist_day <- dist %>% group_by(date, id) %>% 
  dplyr::summarise(mean = mean(dist_release),
                   max = max(dist_release),
                   min = min(dist_release)) %>%
  mutate(date = as_date(date))

# plot average distance from release site per day
ggplot(dist_day)+
  geom_path(aes(date, mean, group=id, color=id))+
  theme_minimal()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  scale_x_date(date_labels = "%b %Y", breaks = "1 month")+
  labs(x= element_blank(), y = "Mean daily distance from release location (m)")+
  scale_color_viridis_d()

# save results
write.csv(dist_day, "distance_release.csv", row.names = FALSE)

```

Total distance moved daily gives an indication of the activity levels of the bird.

```{r distance moved daily}
# read in clean data
data <- read.csv("movebank_20230510_clean.csv") %>% 
  mutate(date_time = as_date(date_time))

# convert to track
track <- make_track(data, .x = utm_easting,.y= utm_northing,
                    .t= date_time, id = id) %>%
  nest(data = -"id")

# loop through distance per day per bird
dates <- unique(data$date)
dist <- data.frame()

birds <- c("Briar", "Nutmeg", "Star", "Aurora", "Robin", "Daisy",
           "Iona", "Sage", "Koda", "Brook", "Clover", "Wobbles",
           "Prem", "Rove", "Valentine", "Marmalade")

for (i in 1:length(birds)){
  for (j in 1:length(dates)){
    
    t <-  track$data[[i]] %>%
      mutate(date = date(t_)) %>%
      filter(date==dates[j])
    
    if (nrow(t)<2){
      next
    }
    
    out <- data.frame(dist= cum_dist(t)/1000,
                      bird = track$id[[i]],
                      date = dates[j])
    
    print(out)
    
    dist <- rbind.data.frame(dist, out)
    
  }}

dist <- dist %>%
  mutate(date = as_date(date))

# save results
write.csv(dist, "distance_daily.csv")

# plot
ggplot(dist)+
  geom_path(aes(date, dist, group = bird, color=bird))+
  theme_minimal()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  scale_x_date(date_labels = "%b %Y", breaks = "1 month")+
  labs(x= element_blank(), y = "Total daily distance moved (km)")+
  scale_color_viridis_d()
```


Can add the weather to the distance moved graph, to visually inspect any potential correlation. The weather data came from the BOM climate data online
http://www.bom.gov.au/climate/data/ observations were drawn from Avalon Airport {station 087113}.

```{r distance and weather}

# read in weather data, which are stored by month
weather <- list.files(pattern = "IDC") %>%
  map_df(~read_csv(., locale=locale(encoding="latin1"),show_col_types = FALSE)) %>%
  mutate(date = as.Date(Date, tryFormats="%d/%m/%Y"))

# read in distance data
dist <- read.csv("distance_daily.csv")

# plot  
ggplot()+
  geom_ribbon(data=weather, aes(x = date, 
                               ymin=`Minimum temperature (°C)`,
                               ymax=`Maximum temperature (°C)`),
              fill="purple", alpha=0.1) + 
  
  xlim(as.Date("2022-10-25"), NA)+
  geom_path(data=dist, aes(date, dist, color=bird))+
  theme_minimal()+
  scale_color_viridis_d()+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
  scale_x_date(date_labels = "%b %Y", breaks = "1 month")+
  labs(x= element_blank(), y = "Total daily distance moved (km)")+
  scale_y_continuous(name = "Temperature (°C)",
    sec.axis = sec_axis( trans=~.*1, name="Distance moved (km)"))

```

### Home range

Core 50% KUD and full 90% KUD home range using kernel utilisation density. 

```{r home range}

# read in clean data
data <- read.csv("movebank_20230510_clean.csv")

# convert GPS fixes to spatial points data frame
locs <- SpatialPointsDataFrame(coordinates(
  cbind(data$utm_easting, data$utm_northing)),
  data = dplyr::select(data, id))

# calculate home range 
## 90% kud
hr_90 <-  kernelUD(locs, h="href", grid=100) %>% 
  getverticeshr(percent = 90)

## 50% kud
hr_50 <-  kernelUD(locs, h="href", grid=100) %>% 
  getverticeshr(percent = 50)

# transform to latlon
## 90% kud
proj4string(hr_90) <- CRS("EPSG:32755") #set crs
hr_90ll <- spTransform(hr_90, CRS("EPSG:4326")) %>% #transform to lat lon
  st_as_sf()

## 50% kud
proj4string(hr_50) <- CRS("EPSG:32755") #set crs
hr_50ll <- spTransform(hr_50, CRS("EPSG:4326")) %>% #transform to lat lon
  st_as_sf()

# background map for plotting
map <- get_map(c(144.435738, -37.900544), zoom=15, maptype = "satellite") 

# plot map
## 90% kud
ggmap(map)+
  geom_sf(data=hr_90ll, aes(fill=id), alpha = .7, inherit.aes = FALSE) +
  scale_fill_viridis_d()

## 50% kud
ggmap(map)+
  geom_sf(data=hr_50ll, aes(fill=id), alpha = .7, inherit.aes = FALSE) +
  scale_fill_viridis_d()

# save results
hr <- data.frame(id = hr_90@data$id,
                 kud90 = hr_90@data$area,
                 kud50 = hr_50@data$area)
write.csv(hr, "hr_area.csv", row.names = FALSE)

```

And daily home range

```{r home range daily}

# read in clean data
data <- read.csv("movebank_20230510_clean.csv") %>%
  mutate(date = as_date(date))

# convert GPS fixes to spatial points data frame
locs <- SpatialPointsDataFrame(coordinates(
  cbind(data$utm_easting, data$utm_northing)),
  data = data)

# loop through birds and days for 90% KUD
birds <- as.character(unique(data$id))
days <- unique(locs[["date"]])

hr_daily <- data.frame()

for (i in 1:length(birds)){
  for (j in 1:length(days)){
    points <- subset(locs, id == birds[i] & date == days[j],
                     select = id)
    if (length(points)<5){
      next
    }
    kud <- kernelUD(points[,1], h="href", grid=1000, extent = 5) %>% 
      getverticeshr(percent = 90) %>%
      st_as_sf() %>%
      mutate(date = days[j])

    print(head(kud, n= 1L))
    
    hr_daily <- rbind.data.frame(hr_daily, kud)
  }}

# extract area per day
hr_area <- st_drop_geometry(hr_daily)
write.csv(hr_area, "hr_daily_area.csv", row.names = TRUE)
    

```

